---
title: "Workflow Pipeline for an Analytical Civic AI Product"
subtitle: "From Toy RAG Demonstration to Marketplace-Ready Municipal Analysis Tool"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- tools::file_path_sans_ext(file_name)



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 

# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```



```{r}
#| label: styling
#| include: false

# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)

```



# <!-- Typst's built-in named colors are the same set as CSS basics: -->
# <!-- black    white    gray     silver -->
# <!-- red      orange   yellow   green -->
# <!-- blue     navy     purple   maroon -->
# <!-- olive    teal     aqua     lime -->
# <!-- fuchsia -->


```{=typst}

#show quote: set text(fill: blue)

```


#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```
\
\

***This is NOT a 'you-can't-do-it' report.  It is a technical assessment of the challenges of building out a fully functioning analytical AI engine as I see it right now.  Every new product has challenges.  If technical challenges cannot be overcome at reasonable cost, consider having AI take it as far as possible, then prepare downloadable files that a human can subject to textual and numerical analysis, independent from the AI engine.***

***See in particular comparison Tables below***

\

# Purpose

This document describes a workflow pipeline suitable for presentation to funders. The goal is to explain the difference between a basic retrieval-based AI demonstration and a serious analytical AI product for municipal transparency, financial analysis, and civic accountability.

The central distinction is this:

> ***A basic RAG chatbot finds and summarizes raw data records.***  

> ***An analytical civic AI must build a trusted evidence layer, a structured data layer, and an auditable reasoning layer before it can safely produce analysis.***


The current proof-of-concept demonstrates that a user can ask questions in plain English and receive source-backed answers from public municipal records. That is useful, and prevalent today, but it is not the same as an analytical product that can reliably calculate growth rates, compare budgets, interpret compensation trends, evaluate assumptions, and connect decisions to later outcomes.


# Product framing

The current proof-of-concept answers who, what and where questions such as:

***"Where in the council packets was compensation discussed?"***

\

The proposed analytical AI should answer effectiveness questions such as:

***"Did the City’s stated justification for compensation increases match the evidence, and did later outcomes support the claim?"***

\
That requires more than retrieval. It requires a system that can:

1. collect municipal documents and data,
2. extract text and tables reliably,
3. convert financial information into validated structured data,
4. preserve citations back to source documents,
5. analyze trends, growth rates, inflation comparisons, and fiscal impacts,
6. identify assumptions, weak evidence, and rhetorical framing,
7. compare claims across years,
8. produce transparent outputs that users can inspect, download, and challenge.


# Current toy RAG compared with proposed analytical AI

| Area | Current toy product / simulated RAG | Proposed analytical AI product |
|---|---|---|
| Core purpose | Retrieve relevant passages from city documents | Analyze claims, numbers, assumptions, trends, trade-offs, and evidence quality |
| Main user question | “Where is this discussed?” | “Is this claim justified by the evidence?” |
| Data type | Mostly unstructured text from PDFs | Unstructured text plus validated structured financial, budget, payroll, staffing, and outcome data |
| Storage | Local files, RDS objects, simulated vector database | Production document store, structured SQL database, vector database, metadata registry, audit logs |
| Chunking | Break PDFs into text chunks | Chunk text, extract tables, tag claims, identify fiscal years, departments, funds, accounts, motions, recommendations |
| Embeddings | Used to find semantically similar passages | Used as one part of retrieval, alongside SQL queries, document metadata, fiscal-year filters, and issue-specific taxonomies |
| Retrieval | Top-k passages returned to the model | Hybrid retrieval: semantic search, keyword search, SQL lookup, metadata filtering, longitudinal linking |
| Citations | Cite retrieved document excerpts | Cite every number, claim, calculation, and interpretation back to source document, table, page, line item, or dataset |
| Numeric work | Minimal or manual | Automated calculations, growth rates, inflation comparisons, variance analysis, fund-level trends, downloadable data frames |
| Financial reliability | Not sufficient for formal analysis | Requires extraction validation, reconciliation, schema rules, data dictionaries, and human review checkpoints |
| Argument analysis | Summary only | Extract claim → evidence → assumption → recommendation → decision → later outcome |
| Evidence quality | Not evaluated | Flags unsupported assertions, cherry-picked comparisons, vendor framing, peer-city claims, and missing methodology |
| Longitudinal memory | Limited to retrieved chunks | Tracks claims, promises, forecasts, votes, outcomes, and later actual results across years |
| User output | Narrative answer with sources | Narrative answer, citations, tables, charts, downloadable datasets, reproducible calculation trail |
| Risk level | Low to moderate | High, because errors in numbers or interpretation can mislead public decision-making |
| Marketplace readiness | Demo / proof of concept | Requires production architecture, quality assurance, governance, validation, security, scalability, legal review, and support model |

# Recommended funder-facing pipeline

## Phase 1: Source acquisition and document registry

The first layer is not AI. It is disciplined public-record collection.

Sources may include:

- council agendas,
- minutes,
- agenda packets,
- budget books,
- annual comprehensive financial reports,
- compensation plans,
- compensation exports obtained by public-information request,
- capital improvement plans,
- staff presentations,
- ordinances and resolutions,
- meeting videos and transcripts (Citizen Portal has made a good start of transcribing video content and indexing it.  Feerally they seem to be current, but they are an entire year behind in transcribing Schertz data making it uninformative for this use case)


***Each source document needs a registry record***

| Field | Purpose |
|---|---|
| Document ID | Permanent internal identifier |
| Source URL | Where the document came from |
| Collection date | When it was captured |
| Document type | Budget, packet, minutes, annual report, payroll file, etc. |
| Meeting date / fiscal year | Time anchor |
| Department / fund / topic | Analytical filtering |
| Version status | Draft, adopted, amended, replaced |
| File hash | Detects whether the file changed |
| Public citation label | Allows source-backed answers |

The first major funding challenge is this:

> ***The AI product is only as reliable as the municipal evidence layer underneath it.***

\

## Phase 2: Text extraction, table extraction, and document parsing

The current toy RAG can tolerate imperfect text extraction. An analytical product cannot.

For simple retrieval, it may be acceptable if a paragraph is mostly correct. For financial analysis, that is not enough. The system must distinguish:

- a table title,
- row labels,
- column headings,
- fiscal years,
- totals,
- subtotals,
- notes,
- amended amounts,
- proposed amounts,
- adopted amounts,
- actual amounts,
- one-time transfers,
- recurring expenditures.

A serious pipeline should separate extraction into two tracks.

| Track | Purpose |
|---|---|
| Text extraction | Used for claims, recommendations, meeting narrative, and staff explanations |
| Table extraction | Used for budgets, compensation, expenditures, revenues, staffing, debt, and capital projects |

This is one of the largest jumps from the current demonstration.

A retrieved paragraph can be close enough for civic search.  A table used for financial analysis must be right.


\
## Phase 3: Structured data normalization

After extraction, the product needs to turn messy municipal records into analysis-ready tables.

For example, a budget table might become:

| fiscal_year | fund | department | account | category | amount_type | amount |
|---|---|---|---|---|---|---:|
| 2020 | General Fund | Police | Salaries | Personnel | Actual | 12400000 |
| 2026 | General Fund | Police | Salaries | Personnel | Proposed | 18900000 |

This is the step that turns PDFs into numbers.

The product needs data dictionaries for terms such as:

- actual,
- adopted,
- amended,
- proposed,
- projected,
- recurring,
- one-time,
- transfer,
- personnel,
- operations,
- capital,
- debt service,
- general fund,
- enterprise fund.

Without this layer, the AI may appear analytical while actually comparing mismatched figures.

> The expensive part is not the chatbot.  
> The expensive part is building a trustworthy municipal data model.

\

## Phase 4: Validation and reconciliation

Before numbers are used by the AI, they need automated and human-auditable checks.

| Validation check | Why it matters |
|---|---|
| Do line items sum to department totals? | Detects extraction errors |
| Do department totals sum to fund totals? | Detects missing rows |
| Do adopted totals match the published budget summary? | Confirms the parsed table agrees with the official document |
| Are fiscal years consistently labeled? | Prevents comparing FY2024 adopted to FY2025 proposed by mistake |
| Are one-time and recurring items separated? | Prevents false claims about structural growth |
| Are transfers double-counted? | Avoids overstating expenditure growth |
| Are employee names, job titles, and salaries normalized? | Enables compensation analysis |
| Are inflation comparisons tied to the correct time period? | Prevents misleading CPI comparisons |

This phase should produce a visible audit trail:

```text
Source document
  ↓
Extracted table
  ↓
Cleaned table
  ↓
Validation checks
  ↓
Approved analytical dataset
```

\

## Phase 5: Analytical database layer

A serious product should not rely on the language model to perform calculations from raw text.

The model should call tools.

The analytical layer should include:

- SQL database for structured municipal data,
- vector database for unstructured document retrieval,
- document store for original PDFs,
- calculation engine for growth rates, inflation adjustment, variance analysis, and trend detection,
- reproducible R or Python scripts for higher-risk analysis,
- downloadable tables for users.

A simplified architecture is:

```text
Public documents and public-information files
        ↓
Document registry
        ↓
Text extraction + table extraction
        ↓
Structured data cleaning
        ↓
Validation and reconciliation
        ↓
Analytical database
        ↓
AI retrieval + SQL + calculation tools
        ↓
Narrative answer + citations + charts + downloadable data
```

The marketplace point is important:

> The model should not be trusted to invent calculations.  
> It should retrieve validated data, run defined calculations, and explain the result.

\

## Phase 6: Claim, evidence, and assumption extraction

This is the most distinctive part of the proposed product.

An analytical AI should not merely summarize staff reports. It should separate claims, evidence, assumptions, recommendations, decisions, and later outcomes.

| Object | Example |
|---|---|
| Claim | “The City must increase salaries to remain competitive.” |
| Evidence | “A peer-city comparison shows Schertz below market.” |
| Assumption | “The selected peer cities are appropriate comparators.” |
| Recommendation | “Adopt a 5% across-the-board raise.” |
| Decision | “Council approved the raise.” |
| Later outcome | “Vacancy rates did or did not improve in the following year.” |

This creates an argument ledger.

That ledger allows the AI to answer accountability questions:

- What evidence was offered?
- Was the evidence internal or external?
- Was the comparator group justified?
- Was the claim repeated in later years?
- Did the promised result occur?
- Did the recommendation match the evidence?
- Was there a less expensive alternative?
- What was not funded because this was funded?

This moves the product from search to governance analysis.

\

## Phase 7: Evidence-quality scoring

The product should not merely repeat official explanations. It should classify the quality of evidence.

| Evidence type | Analytical warning |
|---|---|
| Internal historical data | Usually stronger if definitions are stable |
| Audited financial data | Stronger, but may lag current decisions |
| Public-information payroll data | Strong if complete and reconciled |
| Peer-city comparison | Highly sensitive to comparator selection |
| Vendor study | Possible conflict of interest |
| Staff assertion | May be true, but requires support |
| Anecdote | Weak evidence unless corroborated |
| “Industry standard” claim | Needs source and definition |
| Urgency framing | Requires scrutiny, not automatic acceptance |
| Forecast | Must later be compared to actual results |

Even a simple figure such as “47% growth” embeds choices about:

- start year,
- end year,
- inflation measure,
- fund selection,
- inclusion or exclusion of transfers,
- recurring versus one-time costs,
- department mapping,
- whether population growth is relevant,
- whether service levels changed.

A good analytical AI should expose those choices.

\

## Phase 8: Longitudinal accountability layer

This is where the product becomes meaningfully different from most civic AI tools.

The system should track claims over time, measured by EFFECTIVENESS, not by efficiency.

| Prior claim | Later test |
|---|---|
| “This raise will improve retention.” | Anyone would know in advance that turnover decreases with rise in pay.  Did the action improve effectiveness?  By what method or metric? |
| “This equipment will reduce maintenance costs.” | Did maintenance spending decline?  By what method or metric?|
| “This fee increase is temporary.” | Was it later reversed? By what method or metric? |
| “This position is needed because workload increased.” | Was workload data ever provided?  Did the action improve effectiveness?  By what method or metric?|
| “This project will cost $X.” | What was the final cost? |
| “This is a one-time expenditure.” | Did it become recurring? |

This requires linking across:

- staff reports,
- budget books,
- meeting minutes,
- adopted ordinances,
- later budgets,
- payroll data,
- project reports,
- audit documents.

That is not ordinary RAG. It makes longitudinal data available for study of civic effectiveness.

\

# Marketplace-scale challenges beyond the current demo

## PDF-to-data reliability

The biggest technical challenge is converting PDFs into accurate structured data.

> Retrieval can tolerate imperfect extraction.  
> Financial analysis cannot.

If a chatbot misses a paragraph, the answer may be incomplete. If an analytical AI misreads a table, the answer may be wrong.

> ***The platform GroundX.com has a very promising retrieval system built for superior extraction from pdf files.  They are part of eyelevel.ai.  "APIs for Enterprise-Grade RAG. Built on Kubernetes and fine-tuned open source models. 50% more accurate. Run us in your secure data center."***

\

## Municipal data is inconsistent
\
Cities do not always use the same categories every year.

A department may be renamed. A fund may be reorganized. A line item may move. A cost may shift from operating to capital. A transfer may appear as both revenue and expense.

The product needs a normalization strategy. Otherwise, the AI may compare things that look similar but are not analytically equivalent.

\

## The system must distinguish facts from derived calculations
\
The product should clearly separate raw source facts, derived calculations, contextual comparisons, interpretations, and policy judgments.

| Layer | Example |
|---|---|
| Raw source fact | “FY2020 General Fund expenditures were $42.1 million.” |
| Derived calculation | “The increase from FY2020 to FY2026 was 47%.” |
| Contextual comparison | “That exceeded CPI growth over the same period.” |
| Interpretation | “The increase appears structurally significant.” |
| Policy judgment | “Council should request a recurring-cost explanation.” |

This distinction is crucial for public trust. The AI should never blur these layers.

\

## Methodology can hide premature conclusions
\
A premature conclusion can be introduced through:

- period selection,
- peer-city selection,
- inflation measure,
- fund selection,
- employee group selection,
- treatment of vacancies,
- treatment of benefits,
- inclusion or exclusion of one-time costs.

The product should not pretend methodology is neutral. It should show the methodological choices.

\

## Citations must apply to numbers, not just prose

\
A normal RAG answer cites a paragraph.

An analytical AI must cite:

- the document,
- the page,
- the table,
- the row,
- the column,
- the extracted value,
- the cleaned value,
- the calculation formula.

For funders, this should be understood as a core product requirement, not a nice-to-have.

\

## Human review remains necessary
\
For marketplace use, the system should include review checkpoints.

| Risk level | Example | Review needed |
|---|---|---|
| Low | Find a meeting packet | Minimal |
| Moderate | Summarize staff recommendation | Spot check |
| High | Calculate budget growth | Validation checks |
| Very high | Claim that staff evidence was weak | Human analytical review |
| Very high | Public-facing fiscal risk conclusion | Human review plus methodology disclosure |

This is not a weakness. It is part of responsible product design.

\

## The buyer may not be the natural beneficiary
\
Many civic AI vendors sell to city staff. Their products usually improve service delivery, resident communication, and staff efficiency.

An analytical accountability product benefits:

- council members,
- board members,
- journalists,
- citizens,
- watchdog groups,
- civic researchers,
- candidates,
- advocacy groups,
- internal auditors.

That creates a market challenge:

> ***The entity with the data may not want to buy the tool that makes the data easier to scrutinize.***

This supports the case for an independent civic-transparency product outside city infrastructure.

\

# Proposed product architecture

\

## Layer 1: Evidence collection
\
Purpose: collect and preserve source materials.

Components:

- document scraper,
- public-information file importer,
- PDF archive,
- file hashing,
- metadata registry,
- source URL preservation,
- document-type classifier.

Output:

> ***A trusted public-record library.***

\

## Layer 2: Retrieval layer
\
Purpose: find relevant documents and passages.

Components:

- text extraction,
- chunking,
- embeddings,
- vector database,
- keyword search,
- metadata filters,
- citation labels.

Output:

> ***Source-backed document retrieval.***

This is approximately where the current toy product lives.

\

## Layer 3: Structured data layer
\
Purpose: convert financial and operational records into analysis-ready data.

Components:

- table extraction,
- spreadsheet ingestion,
- data cleaning,
- schema mapping,
- department/fund/account normalization,
- validation checks,
- reconciliation to published totals.

Output:

> ***Trusted municipal data tables.***

\

## Layer 4: Analytical tool layer
\
Purpose: perform calculations outside the language model.

Components:

- SQL query engine,
- R or Python calculation scripts,
- inflation adjustment functions,
- growth-rate functions,
- variance analysis,
- trend analysis,
- control-chart or statistical-process tools where appropriate,
- charting tools,
- data export tools.

Output:

> ***Reproducible calculations and downloadable results.***

\

## Layer 5: Argument and assumption layer
\
Purpose: analyze policy reasoning, not merely numbers.

Components:

- claim extraction,
- recommendation extraction,
- evidence tagging,
- assumption detection,
- rhetorical pattern detection,
- precedent checking,
- longitudinal claim tracking.

Output:

> ***A structured map of municipal decision logic with suggestions for better measures and metrics to assess effectiveness, not just efficiency.***

\

## Layer 6: AI answer layer
\
Purpose: communicate findings clearly.

Components:

- chat interface,
- answer generator,
- citation engine,
- uncertainty warnings,
- calculation explanations,
- chart/table generator,
- downloadable data frames,
- “show your work” panel.

Output:

> ***Public-facing analytical answers with source support.***

\

## Layer 7: Governance and quality assurance layer
\
Purpose: prevent the tool from becoming a confident misinformation machine.

Components:

- audit logs,
- model prompts under version control,
- dataset versioning,
- validation reports,
- human review workflow,
- source confidence scoring,
- red-team testing,
- user feedback loop.

Output:

> ***A product that can be trusted, challenged, and improved.***

\

# Suggested funder-facing development stages

\

## Stage 1: Retrieval MVP
\
Goal:

> ***Demonstrate that users can ask natural-language questions and receive source-backed answers from municipal records.***

This has already been largely achieved in toy form.

Deliverables:

- document ingestion,
- chunking,
- embeddings,
- vector search,
- source citations,
- basic Shiny or chat interface.

Primary limitation:

> ***It finds information but does not reliably analyze numbers.***

> ***The platform GroundX.com has a very promising retrieval system built for superior extraction from pdf files.  They are part of eyelevel.ai.  "APIs for Enterprise-Grade RAG. Built on Kubernetes and fine-tuned open source models. 50% more accurate. Run us in your secure data center."***

\

## Stage 2: Structured-data MVP
\
Goal:

> ***Demonstrate that budget, payroll, and financial tables can be converted into validated data frames.***

Deliverables:

- selected budget books parsed,
- selected payroll files imported,
- data dictionary created,
- fiscal-year schema created,
- validation checks built,
- downloadable tables produced.

Primary risk:

> ***PDF table extraction may require document-specific handling.***



\

## Stage 3: Analytical MVP
\
Goal:

> Allow the AI to answer limited numerical questions using validated data.

Example questions:

- How much did General Fund spending grow from FY2020 to FY2026?
- Which departments accounted for most of the increase?
- How did compensation growth compare with total expenditure growth?
- What was the annualized growth rate?
- How does the increase compare with inflation?

Deliverables:

- SQL or DuckDB analytical layer,
- calculation functions,
- chart outputs,
- downloadable data,
- calculation citations.

Primary risk:

> ***The system must prevent apples-to-oranges comparisons.***

\

## Stage 4: Accountability MVP
\
Goal:

> Link staff claims to evidence, recommendations, decisions, and later outcomes.

Example questions:

- What hard data showing evidence of need was offered for the compensation increase?
- Were peer cities used, and were they justified?
- Did the stated effectiveness problem improve afterward?
- Was this described as one-time or recurring?
- Has the City made the same claim in prior years?

Deliverables:

- claim/evidence extraction,
- assumption tagging,
- longitudinal issue tracking,
- outcome comparison,
- evidence-quality flags.

Primary risk:

> ***This moves from information retrieval into judgment-support, so transparency and human review become essential.***

\

## Stage 5: Marketplace pilot
\
Goal:

> Test the product with real users outside the builder’s personal workflow.

Potential users:

- council members,
- school board members,
- journalists,
- civic watchdogs,
- resident groups,
- candidates,
- public-finance researchers.

Deliverables:

- hosted application,
- role-based user testing,
- feedback collection,
- onboarding materials,
- sample use cases,
- disclaimers,
- source limitations,
- support model.

Primary risk:

> ***The most natural beneficiaries may not be the government entities that control the documents.***

\

# Funder narrative
\
The following language may be used directly in a concept paper or funding pitch.

- ***We have demonstrated the first layer of a civic AI system: retrieval from public municipal records using chunking, embeddings, and source-backed answers. That is useful, but it is not enough for serious public decision support.***

- ***The next stage is to build an analytical AI system that can convert public documents and public-information spreadsheets into validated structured data, connect those data to the original records, and allow users to ask questions about financial growth, compensation, budget changes, assumptions, trade-offs, and evidence quality.***

- ***The challenge is not merely adding a better chatbot. The hard work is building a trustworthy evidence pipeline: document collection, table extraction, data normalization, validation, structured storage, calculation tools, citation tracking, and human-auditable methodology.***

- ***The resulting product would help citizens and oversight bodies move beyond finding documents toward understanding whether public decisions were supported by evidence, whether claims were consistent over time, and whether promised outcomes actually occurred.***

\

# Blunt summary for funders

The current toy product proves that the front door can work.

A user can ask a question. The system can retrieve relevant chunks. The AI can generate a sourced answer.

But a marketplace-ready analytical AI requires several harder things (some are much harder things):

1. Reliable PDF-to-data extraction
2. Validated financial data models
3. SQL, R, or Python calculation tools
4. Citation trails for every number
5. Longitudinal tracking across fiscal years
6. Claim, evidence, and assumption extraction
7. Methodology transparency
8. Human review for high-risk conclusions
9. A sustainable hosting and update pipeline
10. A buyer/user model that works despite institutional resistance

That is the real funding case:

> ***We are not asking for money to build a chatbot.***
> ***We are asking for money to build the public evidence infrastructure that allows AI to safely perform civic analysis.***

\

# Appendix: Practical implementation notes

\

## Likely transition from local proof of concept to enterprise-ready stack
\

A practical prototype runs with the existing R-centered workflow, but a marketplace-ready analytical AI product should not be built around R as the production backbone.

R has been extremely useful for exploration, municipal compensation analysis, charting, Quarto reporting, and rapid proof-of-concept work. It should remain part of the project where it is strongest: statistical exploration, analyst notebooks, reproducible reports, and specialized civic-analysis prototypes.

But the enterprise product needs a different spine:

- Python for ingestion, orchestration, extraction, model integration, testing, and production services.
- SQL/PostgreSQL for validated structured data.
- A production vector database such as Pinecone, Qdrant, Weaviate, or pgvector/Postgres for retrieval.
- Cloud object storage for raw PDFs, extracted text, tables, logs, and versioned artifacts.
- API services for the chatbot, analytical tools, user interface, and data export.
- Formal validation, monitoring, security, and deployment workflows.

The core points for funders:

> ***The current R-based workflow proves the analytical concept.***

> ***An enterprise product requires a production data platform, software-engineering discipline, and auditable AI infrastructure.***

\

### Recommended platform transition
\

| Function | Current proof-of-concept approach | Enterprise-ready direction | Why the transition matters |
|---|---|---|---|
| Document registry | R data frame, CSV, RDS, or DuckDB table | PostgreSQL or managed cloud SQL database | A production registry needs durable IDs, document versioning, audit fields, relational joins, access controls, and reliable backups. |
| Raw PDF storage | Local files and Cloudflare R2 | Cloud object storage such as Cloudflare R2, AWS S3, Azure Blob, or Google Cloud Storage | Raw documents need durable storage, version control, file hashing, lifecycle rules, and repeatable reprocessing. |
| Text extraction | R packages and command-line PDF tools | Python extraction pipeline using tools such as PyMuPDF, pdfplumber, Apache Tika, OCR services, or cloud document AI tools | Enterprise extraction needs scalable batch processing, error handling, page-level outputs, retry logic, and extraction-quality logs. |
| OCR | Manual or ad hoc | OCR service or containerized OCR workflow | Scanned packets, agenda exhibits, and older PDFs require consistent OCR with confidence scoring and human-review flags. |
| Table extraction | R/Python experiments with human validation | Python-based table extraction plus validation workflow; possible use of cloud document AI for hard PDFs | Financial analysis depends on accurate tables. Extraction must preserve row labels, column headers, fiscal years, totals, notes, and source-page references. |
| Structured analytical storage | RDS files, DuckDB, SQLite, or local Postgres | PostgreSQL, warehouse layer, or lakehouse-style architecture | Analytical data needs schemas, constraints, migrations, repeatable loading, query performance, and multi-user access. |
| Local analytics engine | R and DuckDB | SQL-first analytics with Python services; DuckDB retained for analyst sandboxing and fast local testing | DuckDB is excellent for portable analytical work and supports Python and R APIs, but production workflows should have a shared database of record. :contentReference[oaicite:0]{index=0} |
| Vector retrieval | Simulated vector store, local embeddings, or R objects | Pinecone, Qdrant, Weaviate, or pgvector/Postgres | Production retrieval needs indexing, metadata filtering, namespaces, scaling, monitoring, and stable APIs. Pinecone supports hybrid retrieval patterns that combine dense and sparse signals; pgvector brings vector search into PostgreSQL with index options such as HNSW and IVFFlat. :contentReference[oaicite:1]{index=1} |
| Hybrid search | Basic semantic retrieval | Dense vector search plus keyword/BM25 search plus metadata filtering | Civic records often require exact terms, fiscal years, ordinance numbers, fund names, vendor names, and meeting dates. Semantic search alone will miss some of that. Pinecone’s documentation explicitly recommends hybrid approaches because semantic and lexical search have different failure modes. :contentReference[oaicite:2]{index=2} |
| Embeddings | Generated locally or experimentally | Scheduled embedding pipeline with model versioning, batching, retries, and re-embedding policy | Enterprise retrieval needs to know which embedding model created each vector and when vectors must be regenerated. |
| Analytical calculations | R functions and Quarto reports | Tool-called SQL/Python calculation services with tested functions | The language model should not invent calculations. It should call tested tools for growth rates, inflation adjustments, variance analysis, fund comparisons, and downloadable tables. |
| Interface | Shiny or Quarto dashboard | Web application with API backend; Shiny may remain useful for internal prototypes | Shiny is excellent for analyst-built dashboards, but a public SaaS-style product needs authentication, usage logging, role management, API separation, and deployment practices beyond a typical Shiny app. |
| Charts and downloads | R-generated charts, CSV, RDS, XLSX | Server-side chart/data export service producing CSV, XLSX, JSON, and possibly reproducible notebooks | Public users need downloadable results and a visible calculation trail. RDS is useful internally but is not a public-facing interchange format. |
| AI orchestration | Direct calls from R scripts or Shiny | Python service layer using structured tool calls, retrieval services, SQL services, and response validation | The AI layer should route questions to retrieval, SQL, calculation, or refusal paths rather than letting the model improvise. |
| Testing | Manual inspection and Quarto debugging | Unit tests, integration tests, golden-answer tests, extraction tests, retrieval tests, and regression tests | A public analytical AI cannot be trusted unless the pipeline is tested after every update. |
| Deployment | Local machine and shinyapps.io | Containerized services using Docker, CI/CD, cloud hosting, secrets management, and monitoring | Enterprise deployment needs repeatability, uptime, controlled releases, rollback, and log visibility. |
| Governance | Personal review and manual caution | Formal data governance, model governance, source limitations, review tiers, audit logs, and user feedback loops | This product will make claims about public finance and governance. It must show its work and preserve an audit trail. |

\

### Recommended enterprise architecture
\

A realistic architecture would separate the system into services rather than treating the chatbot as one large application.

\


```{=typst}
#set table(
  inset: 5pt,
)
```

| ***Step*** | ***Pipeline layer*** | ***Purpose*** |
|:---:|:---|:---|
| 1 | Public websites, PDFs, spreadsheets, and FOIA files | Source material collected from official public records and legally obtained files |
| 2 | Ingestion service | Downloads, imports, hashes, catalogs, and schedules source updates |
| 3 | Raw document storage | Stores original PDFs, spreadsheets, transcripts, and extracted artifacts |
| 4 | Document registry database | Maintains document IDs, source URLs, meeting dates, fiscal years, document types, and version history |
| 5 | Text extraction, OCR, and table extraction | Converts documents into searchable text and candidate structured tables |
| 6 | Validation and reconciliation pipeline | Checks extracted figures against published totals and flags errors for review |
| 7 | Structured SQL database | Stores validated budget, payroll, staffing, project, and financial data |
| 8 | Vector database / hybrid retrieval index | Supports semantic retrieval, keyword retrieval, and metadata-filtered document search |
| 9 | Python API layer | Provides production services for retrieval, SQL queries, calculations, exports, and AI tool calls |
| 10 | AI orchestration layer | Routes user questions to search, SQL, calculation tools, citation tools, or refusal paths |
| 11 | Web application | Provides the public-facing interface for questions, tables, charts, and downloads |
| 12 | Narrative answers, citations, charts, and downloadable data | Delivers user-facing results with visible sources and calculation trails |

: Recommended enterprise architecture {#tbl-enterprise-architecture tbl-colwidths="[8, 32, 60]"}


\

The important design rules are:

> The AI should not be the database, the calculator, or the source of truth.  

> The AI should be the user-facing reasoning and explanation layer sitting on top of validated data systems.

\

Role of R going forward

\

R should not be abandoned, but its role must change.

R remains valuable for:

- analyst notebooks,
- Quarto reports,
- exploratory compensation analysis,
- chart prototypes,
- statistical review,
- one-off civic investigations,
- validating whether a proposed analytical feature is worth building.


R should not be the primary tool for:

- production ingestion services,
- large-scale API development,
- authentication and user management,
- production AI orchestration,
- scalable document processing,
- enterprise deployment,
- multi-tenant application architecture,
- long-term product operations.

\

A reasonable division of labor is:
\

| Layer                          | Best-fit tool direction                                |
| ------------------------------ | ------------------------------------------------------ |
| Exploration and civic analysis | R, Quarto, DuckDB                                      |
| Production ingestion           | Python                                                 |
| PDF/OCR/table extraction       | Python and document-processing services                |
| Structured storage             | PostgreSQL / cloud SQL                                 |
| Analytical querying            | SQL, Python, possibly DuckDB for sandbox work          |
| Vector retrieval               | Pinecone, Qdrant, Weaviate, or pgvector                |
| AI orchestration               | Python service layer                                   |
| Public application             | Modern web framework plus API backend                  |
| Reporting prototypes           | Quarto and R                                           |
| Enterprise monitoring          | Cloud logging, CI/CD, test suites, observability tools |

\

***Pinecone, pgvector, and vector-platform caution***
\

Pinecone may be attractive because it is a managed vector database designed for production retrieval workloads. It can support hybrid retrieval patterns, where dense semantic search is combined with sparse or lexical signals. That matters because city records are full of exact names, fund labels, ordinance numbers, agenda-item IDs, and fiscal-year references that pure semantic search can miss.

However, Pinecone should not be treated as the whole analytical platform.

It stores and retrieves embedded records. It does not solve:

- PDF extraction,
- financial table validation,
- data normalization,
- SQL analysis,
- inflation calculations,
- audit trails,
- claim/evidence mapping,
- human review,
- public-facing methodology disclosure.

\

***pgvector***

pgvector is another serious option because it allows vector search inside PostgreSQL. That can simplify architecture by keeping structured municipal data, metadata, and embeddings closer together. The trade-off is that the team becomes responsible for more database tuning, indexing, scaling, and operations. pgvector supports approximate-nearest-neighbor index strategies such as HNSW and IVFFlat, each with performance and recall trade-offs.

The platform decision should not be framed as “Pinecone versus SQL.” A serious analytical civic AI likely needs both:

- SQL for validated facts and calculations.
- Vector search for retrieving relevant text passages.
- Keyword search for exact terms.
- Object storage for source documents.
- AI orchestration to decide which tool to use.
- Suggested first enterprise analytical use case

The first serious analytical use case should remain narrow, but it should be implemented in the new architecture rather than extended only inside the R prototype.

***A good candidate would be:***

Explain General Fund expenditure growth over a fixed period, identify the departments and categories contributing most to the increase, cite every source number used in the calculation, and provide the user with a downloadable table of the underlying data.

This use case is strong because it forces the product to solve the hard problems:

- collecting official budget documents,
- extracting relevant financial tables,
- preserving page-level citations,
- mapping fiscal years consistently,
- distinguishing actual, adopted, amended, and proposed figures,
- normalizing departments and categories,
- validating totals against published budget summaries,
- loading cleaned figures into SQL,
- calculating dollar growth, percentage growth, annualized growth, and inflation-adjusted comparisons,
- identifying the largest contributors to the increase,
- explaining methodology in plain English,
- producing downloadable CSV/XLSX/JSON outputs,
- preserving an audit trail from answer back to source document.

It is narrow enough to be feasible but serious enough to demonstrate the difference between a toy chatbot and an analytical civic AI product.

\

***Additional cautions for funders***

The hardest parts of this project are not the visible chatbot features.

The hard parts are:

***Data reliability***

- Municipal PDFs were not designed to become databases. The extraction and validation layer will take serious work.

***Definition control***

- Terms such as “General Fund,” “personnel,” “actual,” “adopted,” “amended,” and “recurring” must be defined consistently.

***Version control***

- Budgets and staff reports may change. The product must know which version it used.

***Citation precision***

- Public trust requires citations not only for paragraphs but also for numbers, table cells, formulas, and derived calculations.

***Hybrid retrieval***

- Semantic search alone is not enough for analyzing civic records. The system must combine vector search, keyword search, metadata filters, and SQL.

***Model discipline***

- The AI must call tools and report results. It should not perform high-stakes calculations from memory or from loosely retrieved text.

***Human in the Loop***

- High-risk outputs, especially those criticizing the quality of official evidence, should have review tiers.

***Cost control***

- Embedding, OCR, vector storage, cloud hosting, monitoring, and LLM calls all create recurring costs.

***Institutional resistance***

- The users who benefit most may be citizens, journalists, candidates, and oversight officials, not necessarily the municipal staff who control the documents.

***Product liability and reputation risk***

- A wrong financial conclusion could damage trust quickly. The product must be conservative, auditable, and transparent.



The R prototype can show what is possible.
The enterprise product must be rebuilt around production data engineering, SQL-backed analytics, vector retrieval, Python services, validation workflows, and public auditability.
